Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-04 paper/paper.tex
writing tweaks and comments

\newcommand{\bang}{~\textbf{!}~}
% Tentatively using uparrow for linear to non-linear to be in line with
% Pfennings adjoint functional programming
\newcommand{\ltonl}[1]{~\uparrow #1}
\newcommand{\nil}{\mathsf{nil}}
\newcommand{\cons}{\mathsf{cons}}
\newcommand{\epscons}{\epsilon\mathsf{cons}}
\newcommand{\data}{\mathsf{data}}
\newcommand{\where}{\mathsf{where}}
\newcommand{\Trace}{\mathsf{Trace}}
\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
\newcommand{\internalize}[1]{\lceil#1\rceil}

@@

To gain intuition for working in \theoryabbv, we study some examples of
grammars and parses of them internal to \theoryabbv. In each of our examples, we will look at
grammars over the three character alphabet $\{ a , b , c\}$.

Each letter of the
alphabet corresponds to a base type in our calculus. For instance, the grammar
for the single character $a$ has a single parse tree for the string \stringquote{a}.
and no parse trees at any other strings.

\paragraph{Finite Grammars}
First consider finite grammars --- those built from base types via disjunctions $\oplus$ and
concatenations $\otimes$. For instance, $(a \otimes b) \oplus c$.
In \cref{fig:fingram} we find a proof tree that constructs a term of type
$(a \otimes b) \oplus c$ in the context $x : a, y : b$. We think of this term as
a parse tree internal to \theoryabbv.

Parses of a fixed string $w$ into a grammar $A$ are internally represented as
terms of type $A$ in the context $\internalize w$, where $\internalize w$ is a
comma-separated list of the constituent characters of $w$. In particular, to
parse \stringquote{ac}, we build parse trees in the context
$\internalize{\stringquote{ab}} = x : a , y : b$. Thus the term
$\inl(x \otimes y)$ from \cref{fig:fingram} details the
manner by which \stringquote{ac} matches the grammar $(a \otimes b) \oplus c$.

@@

@@

\paragraph{Regular Expressions}
Next, we demonstrate how to parse a regular expression in \theoryabbv. Similar
to finite grammars, regular
expressions are built inductively using base types, $\oplus$, and $\otimes$ ;
however, unlike finite grammars, regular expressions additionally have a
restricted form of recursion via Kleene star $(\cdot)^{\ast}$.

@@

For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
$A^{*}$ to be an inductive linear
type, as shown in \cref{fig:kleenestarinductive}. $A^{*}$ has two constructors:
$\nil$, which builds a parse of type $A^{*}$ in the empty context; and $\cons$,
which takes in of $A$ and a parse of $A^{*}$ and linearly combines them to build
a parse of $A^{*}$. The type defined by these constructors belong to the
universe of linear types $L$.

The arrow, $\uparrow$, wrapping these constructors denotes
an interaction between the linear and nonlinear parts of the theory.
In short, a nonlinear term
$M : \ltonl A^{*}$ denotes a linear term in the empty context
$\cdot \vdash M : A^{*}$. Later in
this paper, we will further discuss the behavior of $\uparrow$.
At this moment,
it suffices to read these arrows as allow us to use the reuse the names
``$\nil$'' and ``$\cons$''
as symbols for constructors. That is, the $\nil$ constructor is not linearly
consumed at usage time

Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene
star constructors. In this proof tree we expand some syntactic details that are often
left implicit. For instance, every linear context is parameterized by a nonlinear
context, which is syntactically separated by a semicolon. If the nonlinear
context is empty, we may often omit it; however, in this figure we make this
distinction explicit. Further consider the leaves of the tree. First, the
leaves involving the constructors are nonlinear terms, and thus they occur in
the nonlinear empty context. Then we explicit cast from
an nonlinear $\uparrow$-wrapped type into a linear one. Such casts may often be
made implicitly, such as in $\cref{fig:kleeneabstractproof}$.

@@

The context $(A \otimes A)^{*}$ does not correspond directly to a string, so it is
not quite appropriate
to treat linear terms here as parse \textit{trees}. Instead, it is more
appropriate to think of such terms as parse \textit{transformers}. That is, in
\cref{fig:kleeneabstractproof} the
inhabitant $x : (A \otimes A)^{*}$ in the context is not itself data to be parsed; rather,
$x$ is a parse tree and the term from context $x : (A \otimes A)^{*}$ to the
type $A^{*}$ is a transformation of parse trees of the input grammar into parse
trees of the output grammar.

@@

\paragraph{Non-deterministic Finite Automata}
Automata play a critical
role in defining parsers, as they often serve as an operationalized and
executable presentation of a formal grammar. Therefore, it is crucial that we
can represent them internal to \theoryabbv. As a final example, consider the
representation of $(a^{*} \otimes b) \oplus c$ as a non-deterministic finite
automaton (NFA), shown in \cref{fig:exampleNFA}.

@@

There are two kinds of constructors for a trace: those those modeling the
termination of a trace at a given state and those modeling transitions in the
NFA.\ The constructor $\mathsf{stop}$ denotes the first kind of constructor and
terminates the trace in state 3. The constructor
$\mathsf{2to3}$ denotes the $b$-labeled transition in the NFA from state 2 to state
3. In this case, we build a trace beginning at state 2
by consuming a parse of the label, $b$, and consuming a trace beginning at state 3.
Finally, the constructor $\mathsf{1to2}$ is built quite similarly as $\mathsf{2to3}$;
however, $\mathsf{1to2}$  may transition states without reading a character. This is because the
transition from state 1 to state 2 is labeled by the empty string $\epsilon$, so
transition consumes no input. In this case, when building a trace out of
state 1, it suffices to build a trace out of state 2 \emph{over the same string}.

@@

    \path[->] (1) edge[above] node{$\epsilon$} (2)
              (1) edge[right] node{$c$} (3)
              (2) edge[loop left] node{$a$} (2)
              (2) edge[below] node{$b$} (3);
  \end{tikzpicture}
  \caption{NFA for $(a^{*} \otimes b) \oplus c$}
  \label{fig:exampleNFA}
\end{figure}

@@

%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
\paragraph{Unambiguity}
Ordinarily, we think of a formal grammar as unambiguous when there is at most
one parse tree for any input string; however, this description of unambiguity is
a semantic property that is not expressible in the syntax of \theoryabbv. Rather
than capturing this uniqueness aspect directly, we provide several notions
internal to \theoryabbv that imply unambiguity in the grammars model.

We say that $A$ is \emph{unambiguous} if the unique map $! : A \to \top$ to the
terminal object is a monomorphism. Further, in the presence of distributive coproducts, the following are equivalent:
\begin{itemize}
  \item $A$ is an unambiguous grammar
  \item Given a grammar $B$, all morphisms $B \to A$ are equal
  \item $A \& A \cong A$
\end{itemize}

@@

Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
thus also a monomorphism. That is, $\stringg$ is unambiguous.

@@

\paragraph{Equivalence}
In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
of a formal grammar --- the set of strings accepted by it --- and its strong
generative capacity --- the set of ``structural descriptions'', or parse trees,
that it generates. Two grammars are said to be \emph{weakly equivalent} by
Chomsky if they
agree in their weak generative capacity, and likewise \emph{strongly equivalent}
if they agree in their strong generative capacity. We aim to recover each of
these notions of equivalence internally.

@@

If $f$ and $g$ form an isomorphism, then $A$ and $B$ defined to be \emph{strongly
  equivalent}. This implies Chomsky's notion of strong equivalence, as for any
input string $w$
$A \cong B$ induces a bijective translation of parses $A w \cong B w$.
When abstracting over the latent
structure of formal grammars and describing it in the the language of
categories, it is encouraging to see the same definition of strong equivalence
emerge naturally.

@@

Regular expressions
are among one of the most studied and
well-understood concepts in all of computer science. Pratically, they are
essential tools for lexical analysis and many text
processing tasks. So they are a natural concept to encoding into our
domain-specific grammars language.

Nearly every theoretical discussion of regular expressions is accompanied by a
discussion of finite automata, as it is well known that regular expressions and finite automata are
equivalent as grammar formalisms. In this section, we make the equivalences
between regular expressions and finite automata precise by encoding two theorems
from classical language theory: (1) every regular expression is strongly equivalent
to the traces through some nondeterministic finite automaton (NFA); and, (2) for
every NFA, there is some deterministic finite automaton (DFA) such that the type
of traces through the NFA is weakly equivalent to the type of traces through the DFA.

\newcommand{\states}{\mathsf{states}}
\newcommand{\labelt}{\mathsf{label}}
\newcommand{\transitions}{\mathsf{transitions}}
\newcommand{\epstransitions}{\epsilon\mathsf{transitions}}
\newcommand{\isAcc}{\mathsf{isAcc}}
\newcommand{\init}{\mathsf{init}}
\newcommand{\src}{\mathsf{src}}
\newcommand{\dst}{\mathsf{dst}}
\newcommand{\parse}{\mathsf{parse}}
\newcommand{\print}{\mathsf{print}}
\newcommand{\epssrc}{\epsilon\mathsf{src}}
\newcommand{\epsdst}{\epsilon\mathsf{dst}}

@@

Let the data of an NFA $N$ be given by the following nonlinear data,
\begin{itemize}
\item $N.\states$ --- a finite set of states
\item $N.\init$ --- the initial state
\item $N.\isAcc$ --- a decidable proposition over $N.\states$ denoting which
        states are accepting
\item $N.\transitions$ --- a finite set of labeled transitions
\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling
        the transition
\item $N.\src$ --- a mapping from $N.\transitions$ to source state of a transition
\item $N.\dst$ --- a mapping from $N.\transitions$ to destination state of a transition
\item $N.\epstransitions$ --- a finite set of $\epsilon$-transitions
\item $N.\epssrc$ --- a mapping from $N.\epstransitions$ to source state of a $\epsilon$-transition
\item $N.\epsdst$ --- a mapping from $N.\epstransitions$ to destination state of a $\epsilon$-transition
\end{itemize}

For the above definition of an NFA $N$, in \cref{fig:nfatrace} we define a linear type of traces
through $N$. Traces in $N$ may be built through one of three constructors. We
may terminate a trace at an accepting state with the constructor $\nil$. If we
had a trace beginning at the destination state of a transition, then we may use
the $\cons$ constructor to linearly combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the transition.
Finally, if we had a trace beginning at the destination of an
$\epsilon$-transition then we may use $\epscons$ to pull it back along the
$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.

\begin{figure}
\begin{align*}
\data &~\Trace_{N} : (s : N.\states) \to L~\where\\
      & \nil : \ltonl {\left( \LinPiTy {s} {N.\states} {\LinPiTy {\mathsf{acc}} {N.\isAcc(s)} {\Trace_{N}~s}} \right)}  \\
      & \cons : \ltonl {\left( \LinPiTy {t} {N.\transitions} {(N.\labelt~t) \lto \Trace_{N}~(N.\dst(t)) \lto \Trace_{N}(N.\src(t))} \right)} \\
      & \epscons : \ltonl {\left( \LinPiTy {t} {N.\epstransitions} {\Trace_{N}~(N.\epsdst(t)) \lto \Trace_{N}(N.\epssrc(t))} \right)}
\end{align*}
\caption{Traces through the NFA $N$ as an indexed inductive type}
\label{fig:nfatrace}
\end{figure}

@@

\newcommand{\Agda}{Formalized in Agda}
\begin{theorem}[Determinization, \Agda]
  Given an NFA $N$, there exists a DFA $D$ such that
  $\Trace_{N}~N.\mathsf{init}$ is weakly equivalent to $\Trace_{D}~D.\mathsf{init}$.
\end{theorem}

\begin{proof}
  We present a variant of the Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
  the DFA $D$. Define the
  states of $D$ to be the $\mathbb{P}_{\epsilon}(N.\mathsf{states})$ --- the type of $\epsilon$-closed subsets of
  $N.\mathsf{states}$. A subset is accepting in $D$ if it contains an accepting state
  from $N$. Construct the initial state of $D$ as the
  $\epsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
  of $D$ to send the subset $X$ under the character $c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labelled with the character $c$.

@@

If the NFA trace is a labeled transition $s \overset{c}{\to} s'$ such that
$s \in X$, then we first decompose the trace to a parse of the character $c$ and a
trace starting from $s'$ --- that is, to a term of type
$c \otimes \Trace_{N}(s')$. We then inductively call $f$ on traces out of $s'$
to get traces in a parse of $c$ concatenated with a family of traces
parametrized by $\epsilon$-closed subsets containing $s'$.
. We instantiate this family at $\delta_{c}(X)$, as
$s'$ is reachable from $s$ by a single transition labelled by $c$, so
$s' \in \overline{\epsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
linearly combine the parse of character $c$ and the trace out of $D.\delta_{c}(X)$
to get a trace in $D$ rooted at $X$.

@@

Lastly, if the NFA trace was an $\epsilon$-transition
$s \overset{\epsilon}{\to} s'$ with $s \in X$, then the construction is quite similar. In this
case, we again inductively call $f$ on traces out of $s'$ except we instantiate
the choice of subset containing $s'$ with $X$ itself. We may do this as all
states in $D$ are taken to be $\epsilon$-closed, therefore $s' \in X$ because
$s$ is in $X$ and $\epsilon$-transitions to $s'$.

@@

We define $g$ by induction on traces through $D$. First if the trace stops at
some accepting $\epsilon$-closed subset $X$, then we may choose some state
$s \in X$ such that $s$ is accepting and terminate the trace in $N$ at $s$.

Next if the trace in $D$ is defined via transition
$X \overset{c}{\to} D.\delta_{c}(X)$ then we may choose some state in
$s \in D.\delta_{c}(X)$. Inductively, $g$ sends the trace rooted at
$D.\delta_{c}(X)$ to one rooted at $s$. By definition of $D.\delta_{c}(X)$,
there exists some $s' \in X$ such that $s'$ tranitions to $s$ by some number of
$\epsilon$-transitions followed by a transition labeled by $c$.
We order all
such traces of this form and choose the smallest one, which is then linearly
combined with the parse of $c$ to build an $N$ trace. While using
$N.\mathsf{cons}$ to build this final trace, we appropriately shift the indexing
data to $s'$, which is adherent to the defintion of $D.\delta_{c}(X)$. This concludes the
definition of $g$.

@@

\begin{proof}
In this proof, we present a variant of Thompson's construction
\cite{thompsonProgrammingTechniquesRegular1968} in which we induct on the
structure of $R$ to construct strongly equivalent NFAs. We will sketch this
proof by walking through some of the inductive cases, the remaining cases follow
similar reasoning and the details may be found in the formalized artifact.

@@

Now consider the case where $R$ is the disjunction of two regular expressions
$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
$N_{2}$, and a new marked initial state. The accepting states are precisely
those that were accepting in the subautomata. The initial state has an
$\epsilon$-transition to each of the initial states of the subautomata. Once in
$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
$\epsilon$-transition out of the initial state to take. A schematic
representation of this automaton is given in \cref{fig:disjunctionnfa}.

@@

    \path[->] (init) edge[below] node{$\epsilon$} (init1)
              (init) edge[above] node{$\epsilon$} (init2)
              (init1) edge[above] node{} (dots1)
              (init2) edge[above] node{} (dots2)
              (dots1) edge[above] node{} (acc1)
              (dots2) edge[above] node{} (acc2);
  \end{tikzpicture}
  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
  \label{fig:disjunctionnfa}
\end{figure}

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
\newcommand{\balanced}{\mathsf{balanced}}
\newcommand{\Dyck}{\mathsf{Dyck}}
Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
grammars (CFGs). In many ways, the smallest nontrivial CFG is the Dyck grammar
$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
cut to the essence of what it means to be context free.

@@

\paragraph{Linearity with Dependent Types}
\steven{I don't know if it worth discussing the different influences of
  categorical model here? We take Neel's model and add to it as necessary}

>>>>

\newcommand{\bang}{~\textbf{!}~}
% Tentatively using uparrow for linear to non-linear to be in line with
% Pfennings adjoint functional programming
\newcommand{\ltonl}[1]{~\uparrow #1}
\newcommand{\nil}{\mathsf{nil}}
\newcommand{\cons}{\mathsf{cons}}
\newcommand{\epscons}{\varepsilon\mathsf{cons}}
\newcommand{\data}{\mathsf{data}}
\newcommand{\where}{\mathsf{where}}
\newcommand{\Trace}{\mathsf{Trace}}
\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
\newcommand{\internalize}[1]{\lceil#1\rceil}

@@

To gain intuition for working in \theoryabbv, we study some examples of
grammars and parses of them internal to \theoryabbv. In each of our examples, we will look at
grammars over the three character alphabet $\{ a , b , c\}$. Each letter of the
alphabet is a base type in our calculus. For instance, the grammar
for the single character $a$ has a single parse tree for the string \stringquote{a},
and no parse trees at any other strings.

\paragraph{Finite Grammars}
\max{epsilon? bottom?}
First consider finite grammars --- those built from base types via disjunctions $\oplus$ and
concatenations $\otimes$. For instance, $(a \otimes b) \oplus c$.
In \cref{fig:fingram} we find a proof tree that constructs a term of type
$(a \otimes b) \oplus c$ in the context $x : a, y : b$. We think of this term as
a parse tree internal to \theoryabbv.

For a grammar $A$, a parse tree of a string $w$ can be represented as
a term of type $A$ in the context $\internalize w$, where
$\internalize w$ is a context with one variable for each character of
$w$. For example, to define a parse tree for \stringquote{ab}, we use
the context $\internalize{\stringquote{ab}} = x : a , y : b$. Then
\cref{fig:fingram} shows a typing derivation that shows that $\inl(x
\otimes y)$ defines a valid parse tree for the grammar $(a \otimes b)
\oplus c$.

@@

\max{I think concrete examples here would be helpful}
For this interpretation of terms as parse trees to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
a type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once.
%
Finally, exchange allows for variables in the context to be reordered,
but the ordering of characters in a string cannot be ignored in
parsing: otherwise we would have that $\inl(x\otimes y)$ is a valid
term in context $y:b,x:a$, but there should be no parses of $ba$ in
this grammar.

@@

\paragraph{Regular Expressions}
Regular expressions can be encoded as types using base types,
$\oplus$, and $\otimes$ and the Kleene star $(\cdot)^{\ast}$.  For a
grammar $A$, we can define the Kleene star $A^{*}$ as a particular
\emph{inductive linear type} of linear lists, as shown in
\cref{fig:kleenestarinductive}. Here $A^{*} : L$ means we are defining
a \emph{linear} type. $A^{*}$ has two constructors: $\nil$, which
builds a parse of type $A^{*}$ from nothing; and $\cons$, which
linearly consumes a parse of $A$ and a parse of $A^{*}$ and builds a
parse of $A^{*}$. This linear consumption is defined by the linear
function type $\ltonl$. The arrow, $\uparrow$, wrapping these
constructors means that the constructors by themselves do not consume
any resources, and so are \emph{non-linear} values themselves.

@@

In \cref{fig:kleenestarderivation}, we demonstrate that
\stringquote{ab} matches $(a^{*} \otimes b) \oplus c$ through repeated
application of the Kleene star constructors. In this proof tree we
expand some syntactic details that are often left implicit. For
instance, every linear context is parameterized by a nonlinear
context, which is syntactically separated by a semicolon. If the
nonlinear context is empty, we may often omit it; however, in this
figure we make this distinction explicit. Further consider the leaves
of the tree. First, the leaves involving the constructors are
nonlinear terms, and thus they occur in the nonlinear empty
context. Then we explicit cast from an nonlinear $\uparrow$-wrapped
type into a linear one. Such casts may often be made implicitly, such
as in $\cref{fig:kleeneabstractproof}$.

@@

The context $(A \otimes A)^{*}$ does not correspond directly to a string, so it is
not quite appropriate
to treat linear terms here as parse \textit{trees}. Instead, it is more
appropriate to think of such terms as parse \textit{transformers}. That is, in
\cref{fig:kleeneabstractproof} the
inhabitant $x : (A \otimes A)^{*}$ in the context is not itself data to be parsed; rather,
$x$ is a parse tree and the term from context $x : (A \otimes A)^{*}$ to the
type $A^{*}$ is a transformation of parse trees of the input grammar into parse
trees of the output grammar.
We also see here why $\cons$ and $\nil$ should themselves be
non-linear values: we should be able to use them in constructing multiple parse trees.

@@

\paragraph{Non-deterministic Finite Automata}
\max{A lot of passive voice and indirect language in these paragraphs ``Often one wants to...'' ``Our method of representing...is''.}
Automata play a critical role in defining parsers, as they often serve
as an operationalized and executable presentation of a formal
grammar. As a final example, consider the representation of $(a^{*}
\otimes b) \oplus c$ as a non-deterministic finite automaton (NFA),
shown in \cref{fig:exampleNFA}.

@@

There are two kinds of constructors for a trace: those those modeling the
termination of a trace at a given state and those modeling transitions in the
NFA.\ The constructor $\mathsf{stop}$ denotes the first kind of constructor and
terminates the trace in state 3. The constructor
$\mathsf{2to3}$ denotes the $b$-labeled transition in the NFA from state 2 to state
3. In this case, we build a trace beginning at state 2
by consuming a parse of the label, $b$, and consuming a trace beginning at state 3.
Finally, the constructor $\mathsf{1to2}$ is built quite similarly as $\mathsf{2to3}$;
however, $\mathsf{1to2}$  may transition states without reading a character. This is because the
transition from state 1 to state 2 is labeled by the empty string $\varepsilon$, so
transition consumes no input. In this case, when building a trace out of
state 1, it suffices to build a trace out of state 2 \emph{over the same string}.

@@

    \path[->] (1) edge[above] node{$\varepsilon$} (2)
              (1) edge[right] node{$c$} (3)
              (2) edge[loop left] node{$a$} (2)
              (2) edge[below] node{$b$} (3);
  \end{tikzpicture}
  \caption{NFA for $(a^{*} \otimes b) \oplus c$}
  \label{fig:exampleNFA}
\end{figure}

@@

%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
\paragraph{Unambiguity}

Ordinarily, we think of a formal grammar as unambiguous when there is at most
one parse tree for any input string; however, this description of unambiguity is
a semantic property that is not expressible in the syntax of \theoryabbv. Rather
than capturing this uniqueness aspect directly, we provide several notions
internal to \theoryabbv that imply unambiguity in the grammars model.

\max{what's a terminal object and what's a monomorphism and what's a distributive coproduct}
We say that $A$ is \emph{unambiguous} if the unique map $! : A \to \top$ to the
terminal object is a monomorphism. Further, in the presence of distributive coproducts, the following are equivalent:
\begin{itemize}
  \item $A$ is an unambiguous grammar
  \item Given a grammar $B$, all morphisms $B \to A$ are equal\max{is this internal statement? What is a ``morphism''}
  \item $A \& A \cong A$\max{what is $\cong$?}
\end{itemize}

@@

Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
thus also a monomorphism. That is, $\stringg$ is unambiguous.
\max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}

@@

\paragraph{Equivalence}
In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
of a formal grammar --- the set of strings accepted by it --- and its strong
generative capacity --- the set of ``structural descriptions'', or parse trees,
that it generates. Two grammars are said to be \emph{weakly equivalent} by
Chomsky\max{you can drop the ``by Chomsky'' at this point} if they
agree in their weak generative capacity, and likewise \emph{strongly equivalent}
if they agree in their strong generative capacity. We aim to recover each of
these notions of equivalence internally.

@@

\max{indirect language}
If $f$ and $g$ form an isomorphism, then $A$ and $B$ defined to be \emph{strongly
  equivalent}. This implies Chomsky's notion of strong equivalence, as for any
input string $w$
$A \cong B$ induces a bijective translation of parses $A w \cong B w$.
When abstracting over the latent
structure of formal grammars and describing it in the the language of
categories, it is encouraging to see the same definition of strong equivalence
emerge naturally.

@@

\max{this paragraph isn't really saying anything}
Regular expressions
are among one of the most studied and
well-understood concepts in all of computer science. Pratically, they are
essential tools for lexical analysis and many text
processing tasks. So they are a natural concept to encoding into our
domain-specific grammars language.

\max{who cares what every discussion of regular expressions is accompanied by}
Nearly every theoretical discussion of regular expressions is accompanied by a
discussion of finite automata, as it is well known that regular expressions and finite automata are
equivalent as grammar formalisms. In this section, we make the equivalences
between regular expressions and finite automata precise by encoding two theorems
from classical language theory: (1) every regular expression is strongly equivalent
to the traces through some nondeterministic finite automaton (NFA); and, (2) for
every NFA, there is some deterministic finite automaton (DFA) such that the type
of traces through the NFA is weakly equivalent to the type of traces through the DFA.

\newcommand{\states}{\mathsf{states}}
\newcommand{\labelt}{\mathsf{label}}
\newcommand{\transitions}{\mathsf{transitions}}
\newcommand{\epstransitions}{\varepsilon\mathsf{transitions}}
\newcommand{\isAcc}{\mathsf{isAcc}}
\newcommand{\init}{\mathsf{init}}
\newcommand{\src}{\mathsf{src}}
\newcommand{\dst}{\mathsf{dst}}
\newcommand{\parse}{\mathsf{parse}}
\newcommand{\print}{\mathsf{print}}
\newcommand{\epssrc}{\varepsilon\mathsf{src}}
\newcommand{\epsdst}{\varepsilon\mathsf{dst}}

@@

Let the data of an NFA $N$ be given by the following nonlinear data,
\begin{itemize}
\item $N.\states$ --- a finite set of states
\item $N.\init$ --- the initial state
\item $N.\isAcc$ --- a decidable proposition over $N.\states$ denoting which
        states are accepting
\item $N.\transitions$ --- a finite set of labeled transitions
\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling the transition
\item $N.\src$ and $N.\dst$ --- mappings from $N.\transitions$ to source and target states of a labeled transition
\item $N.\epstransitions$ --- a finite set of $\varepsilon$-transitions
\item $N.\epssrc$ and $N.\epsdst$ --- mappings from $N.\epstransitions$ to source and target states of an $\varepsilon$-transition
\end{itemize}

For the above definition of an NFA $N$, in \cref{fig:nfatrace} we define a linear type of traces
through $N$. Traces in $N$ may be built through one of three constructors. We
may terminate a trace at an accepting state with the constructor $\nil$. If we
had a trace beginning at the destination state of a transition, then we may use
the $\cons$ constructor to linearly combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the transition.
Finally, if we had a trace beginning at the destination of an
$\varepsilon$-transition then we may use $\epscons$ to pull it back along the
$\varepsilon$-transition and construct a trace beginning at the source of the $\varepsilon$-transition.

\begin{figure}
\begin{align*}
\data &~\Trace_{N} : (s : N.\states) \to L~\where\\
      & \nil : \ltonl {\left( \LinPiTy {s} {N.\states} {\LinPiTy {\mathsf{acc}} {N.\isAcc(s)} {\Trace_{N}~s}} \right)}  \\
      & \cons : \ltonl {\left( \LinPiTy {t} {N.\transitions} {(N.\labelt~t) \lto \Trace_{N}~(N.\dst(t)) \lto \Trace_{N}(N.\src(t))} \right)} \\
      & \epscons : \ltonl {\left( \LinPiTy {t} {N.\epstransitions} {\Trace_{N}~(N.\epsdst(t)) \lto \Trace_{N}(N.\epssrc(t))} \right)}
\end{align*}
\caption{Traces through an NFA $N$ as an indexed inductive type}
\label{fig:nfatrace}
\end{figure}

@@

\max{I've seen people use a checkmark or a cute little agda logo for this}
\newcommand{\Agda}{Formalized in Agda}
\begin{theorem}[Determinization, \Agda]
  Given an NFA $N$, there exists a DFA $D$ such that
  $\Trace_{N}~N.\mathsf{init}$ is weakly equivalent to $\Trace_{D}~D.\mathsf{init}$.
\end{theorem}

\begin{proof}
  We present a variant of the Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
  the DFA $D$. Define the
  states of $D$ to be the $\mathbb{P}_{\varepsilon}(N.\mathsf{states})$ --- the type of $\varepsilon$-closed subsets of
  $N.\mathsf{states}$. A subset is accepting in $D$ if it contains an accepting state
  from $N$. Construct the initial state of $D$ as the
  $\varepsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
  of $D$ to send the subset $X$ under the character $c$ to the
  $\varepsilon$-closure of all the states reachable from $X$ via a transition
  labelled with the character $c$.

@@

If the NFA trace is a labeled transition $s \overset{c}{\to} s'$ such that
$s \in X$, then we first decompose the trace to a parse of the character $c$ and a
trace starting from $s'$ --- that is, to a term of type
$c \otimes \Trace_{N}(s')$. We then inductively call $f$ on traces out of $s'$
to get traces in a parse of $c$ concatenated with a family of traces
parametrized by $\varepsilon$-closed subsets containing $s'$.
. We instantiate this family at $\delta_{c}(X)$, as
$s'$ is reachable from $s$ by a single transition labelled by $c$, so
$s' \in \overline{\varepsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
linearly combine the parse of character $c$ and the trace out of $D.\delta_{c}(X)$
to get a trace in $D$ rooted at $X$.

@@

Lastly, if the NFA trace was an $\varepsilon$-transition
$s \overset{\varepsilon}{\to} s'$ with $s \in X$, then the construction is quite similar. In this
case, we again inductively call $f$ on traces out of $s'$ except we instantiate
the choice of subset containing $s'$ with $X$ itself. We may do this as all
states in $D$ are taken to be $\varepsilon$-closed, therefore $s' \in X$ because
$s$ is in $X$ and $\varepsilon$-transitions to $s'$.

@@

We define $g$ by induction on traces through $D$. First if the trace stops at
some accepting $\varepsilon$-closed subset $X$, then we may choose some state
$s \in X$ such that $s$ is accepting and terminate the trace in $N$ at $s$.

Next if the trace in $D$ is defined via transition
$X \overset{c}{\to} D.\delta_{c}(X)$ then we may choose some state in
$s \in D.\delta_{c}(X)$. Inductively, $g$ sends the trace rooted at
$D.\delta_{c}(X)$ to one rooted at $s$. By definition of $D.\delta_{c}(X)$,
there exists some $s' \in X$ such that $s'$ tranitions to $s$ by some number of
$\varepsilon$-transitions followed by a transition labeled by $c$.
We order all
such traces of this form and choose the smallest one, which is then linearly
combined with the parse of $c$ to build an $N$ trace. While using
$N.\mathsf{cons}$ to build this final trace, we appropriately shift the indexing
data to $s'$, which is adherent to the defintion of $D.\delta_{c}(X)$. This concludes the
definition of $g$.

@@

\begin{proof}
In this proof, we present a variant of Thompson's construction
\cite{thompsonProgrammingTechniquesRegular1968} in which we \max{I induct this sentence into the hall of shame for misuse of the word ``induct''}induct on the
structure of $R$ to construct strongly equivalent NFAs. We will sketch this
proof by walking through some of the inductive cases, the remaining cases follow
similar reasoning and the details may be found in the formalized artifact.

@@

Now consider the case where $R$ is the disjunction of two regular expressions
$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
$N_{2}$, and a new marked initial state. The accepting states are precisely
those that were accepting in the subautomata. The initial state has an
$\varepsilon$-transition to each of the initial states of the subautomata. Once in
$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
$\varepsilon$-transition out of the initial state to take. A schematic
representation of this automaton is given in \cref{fig:disjunctionnfa}.

@@

    \path[->] (init) edge[below] node{$\varepsilon$} (init1)
              (init) edge[above] node{$\varepsilon$} (init2)
              (init1) edge[above] node{} (dots1)
              (init2) edge[above] node{} (dots2)
              (dots1) edge[above] node{} (acc1)
              (dots2) edge[above] node{} (acc2);
  \end{tikzpicture}
  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
  \label{fig:disjunctionnfa}
\end{figure}

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
\newcommand{\balanced}{\mathsf{balanced}}
\newcommand{\Dyck}{\mathsf{Dyck}}
Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
grammars (CFGs). \max{In many ways?}In many ways, the smallest nontrivial CFG is the Dyck grammar
$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
cut to the essence of what it means to be context free.

@@

%%    - verified imperative implementation: Lambek logic ala separation
%%      logic?
\paragraph{Relation to Separation Logic}
